---
title: "Lab_Athena_Lab3"
output: html_document
---

### Student Information

Name: Athena Nicole R. Salve

Section: B2

Date: September 2, 2026

GitHub Repository URL: <https://github.com/arsalve28/BIO118>

# 3A Tidying Data Using tidyr

## Introduction

Biological data are often collected and stored in formats that are convenient for recording observations but not necessarily convenient for analysis. For example, information from different variables may be combined in a single column, measurements may be arranged across multiple columns, or some observations may contain missing values. Before statistical analysis or visualization, these data often need to be reorganized into a consistent and analysis-ready structure.

The tidyr package in R provides functions for organizing and reshaping data. A tidy dataset generally follows three principles: each variable is represented by a column, each observation is represented by a row, and each cell contains a single value.

In this exercise, you will practice using selected tidyr functions to reorganize and clean biological datasets. You will work with both small examples and a simulated gene-expression dataset to understand how these techniques can be applied to biological data.

### Practice

## Simulated Gene Expression Dataset

```{r}
# Load the package
library(tidyr)

# Create the simulated gene expression dataset
gene_expression <- data.frame(
  gene_id = c("Gene1", "Gene2", "Gene3", "Gene4"),
  ctrl_brain_rep1 = c(12, 18, 23, 9),
  ctrl_brain_rep2 = c(11, 19, 24, 10),
  treat_brain_rep1 = c(15, 21, 28, 11),
  treat_brain_rep2 = c(14, 22, 30, 12),
  ctrl_liver_rep1 = c(22, 30, 35, 20),
  ctrl_liver_rep2 = c(21, 31, 36, 21),
  treat_liver_rep1 = c(26, NA, 39, 25),
  treat_liver_rep2 = c(27, 34, 40, 26)
)

print(gene_expression)

# Convert the dataset from wide to long format
long_gene_expression <- pivot_longer(
  gene_expression,
  cols = c(
    ctrl_brain_rep1, ctrl_brain_rep2,
    treat_brain_rep1, treat_brain_rep2,
    ctrl_liver_rep1, ctrl_liver_rep2,
    treat_liver_rep1, treat_liver_rep2
  ),
  names_to = "measurement",
  values_to = "expression"
)

# Examine the long dataset
print(long_gene_expression)

# Split measurement into condition, tissue, and replicate
tidy_gene_expression <- separate(
  long_gene_expression,
  measurement,
  into = c("condition", "tissue", "replicate"),
  sep = "_"
)

# Examine the tidy dataset
print(tidy_gene_expression)
```

- The missing expression value belongs to Gene2 under the treatment condition (treat), in liver tissue, for replicate 1 (rep1). I would retain this observation because the reason for the missing value is unknown. It can be handled later based on the requirements of the analysis.

```{r}
# Recreate the tidy dataset using a single piped workflow
tidy_gene_expression <- gene_expression %>%
  pivot_longer(
    cols = c(
      ctrl_brain_rep1, ctrl_brain_rep2,
      treat_brain_rep1, treat_brain_rep2,
      ctrl_liver_rep1, ctrl_liver_rep2,
      treat_liver_rep1, treat_liver_rep2
    ),
    names_to = "measurement",
    values_to = "expression"
  ) %>%
  separate(
    measurement,
    into = c("condition", "tissue", "replicate"),
    sep = "_"
  )

print(tidy_gene_expression)
```

# 3B Transforming Data

## Introduction

Biological datasets often need to be transformed, summarized, and reorganized before they can be visualized or analyzed statistically. The dplyr package in R provides functions that make these tasks easier and more reproducible. These functions can be used to select variables, filter observations, create new variables, arrange data, summarize measurements, and combine information from different datasets. The tidyr package in R provides functions for organizing and reshaping data. A tidy dataset generally follows three principles: each variable is represented by a column, each observation is represented by a row, and each cell contains a single value.

In this exercise, you will work with simulated biological datasets to practice common data-transformation tasks using dplyr. You will also learn how to combine multiple operations into a single workflow using the pipe operator (%>%). These skills will help you prepare biological data for visualization and, later in the course, for statistical analysis.

### Practice

## Simulated Abundance Dataset

```{r}
# Load the package
library(dplyr)

# Create the simulated dataset
set.seed(123)

species_list <- c("Species_A", "Species_B", "Species_C", "Species_D")
region_list <- c("Region_1", "Region_2", "Region_3")

years <- sample(2000:2020, 100, replace = TRUE)
abundance <- sample(1:500, 100, replace = TRUE)
temperature <- sample(15:35, 100, replace = TRUE)

abundance_data <- data.frame(
  species = sample(species_list, 100, replace = TRUE),
  region = sample(region_list, 100, replace = TRUE),
  year = years,
  abundance = abundance,
  temperature = temperature
)

print(abundance_data)

# Filter observations with abundance greater than 100 and temperature above 20 degrees Celsius
filtered_data <- filter(
  abundance_data,
  abundance > 100,
  temperature > 20
)

print(filtered_data)

# Select only species, region, and abundance
selected_data <- select(
  abundance_data,
  species, region, abundance
)

print(selected_data)

# Arrange species alphabetically, then abundance in descending order
arranged_data <- arrange(
  abundance_data,
  species,
  desc(abundance)
)

print(arranged_data)

# Calculate overall total abundance and average temperature
summary_data <- summarize(
  abundance_data,
  total_abundance = mean(abundance) * n(),
  mean_temperature = mean(temperature)
)

print(summary_data)

# Calculate mean abundance for each species
species_summary <- abundance_data %>%
  group_by(species) %>%
  summarize(mean_abundance = mean(abundance))

print(species_summary)

# Use a single pipeline to filter years after 2010, group by region, calculate means, and arrange the results
processed_abundance_data <- abundance_data %>%
  filter(year > 2010) %>%
  group_by(region) %>%
  summarize(
    mean_abundance = mean(abundance),
    mean_temperature = mean(temperature)
  ) %>%
  arrange(desc(mean_abundance))

print(processed_abundance_data)
```

# 3C Wrangling Data Using stringr, forcats, and lubridate

## Introduction

Biological datasets often contain information that needs to be cleaned, standardized, or converted into an appropriate format before analysis.  Species names, gene identifiers, experimental conditions, categorical variables, and dates and times may be recorded inconsistently or stored as text even when they represent structured information.

In this exercise, you will use functions from the stringr, forcats, and lubridate packages to clean and transform biological data. You will practice manipulating character strings, organizing categorical variables, and converting and extracting information from dates and times. These skills are important when preparing biological datasets for visualization and statistical analysis.

### Practice

## Simulated Cell Biology Experiment Data

```{r}
# Load the packages
library(dplyr)
library(stringr)
library(forcats)
library(lubridate)

# Create the simulated cell biology dataset
cell_data <- data.frame(
  cell_line = c("HEK293 ", "HeLa", "A549 ", "HepG2", "HeLa", "HEK293"),
  gene_expression = c(
    "BRCA1_001", "TP53_002", "MYC_003",
    "EGFR_001", "BRCA2_004", "EGFR_005"
  ),
  treatment = c(
    "Control", "control", "Treatment",
    "TREATMENT", "Control", "control"
  ),
  observation_time = c(
    "2023-07-15 08:30:00",
    "2023-07-15 10:45:00",
    "2023-07-16 09:00:00",
    "2023-07-16 11:20:00",
    "2023-07-17 12:15:00",
    "2023-07-17 13:00:00"
  )
)

print(cell_data)

# Remove leading and trailing whitespace from cell_line
cell_data <- cell_data %>%
  mutate(cell_line = str_trim(cell_line))

print(cell_data)

# Standardize treatment values to lowercase
cell_data <- cell_data %>%
  mutate(treatment = str_to_lower(treatment))

print(cell_data)

# Convert treatment and cell_line to factors
cell_data <- cell_data %>%
  mutate(
    treatment = factor(treatment),
    cell_line = factor(cell_line)
  )

levels(cell_data$treatment)
levels(cell_data$cell_line)

# Reorder treatment levels so treatment comes before control
cell_data <- cell_data %>%
  mutate(treatment = fct_relevel(treatment, "treatment", "control"))

levels(cell_data$treatment)

# Reorder cell_line levels by frequency
cell_data <- cell_data %>%
  mutate(cell_line = fct_infreq(cell_line))

levels(cell_data$cell_line)

# Convert observation_time to a datetime object
cell_data <- cell_data %>%
  mutate(observation_time = ymd_hms(observation_time))

print(cell_data)

# Verify the data type of observation_time
class(cell_data$observation_time)

# Extract year, month, day, and hour
cell_data <- cell_data %>%
  mutate(
    year = year(observation_time),
    month = month(observation_time),
    day = day(observation_time),
    hour = hour(observation_time)
  )

print(cell_data)

# Calculate hours since the first observation
cell_data <- cell_data %>%
  mutate(
    hours_since_first_observation = as.numeric(
      difftime(
        observation_time,
        min(observation_time),
        units = "hours"
      )
    )
  )

print(cell_data)
```


# 3D Organizing Files within an R Project

## Introduction

In biological data analysis, data, scripts, and analysis outputs are often stored as separate files. Keeping these files organized makes an analysis easier to manage, reproduce, and share.

In this exercise, you will continue working within your existing BIO118 R Project and practice organizing files associated with a small data analysis. You will create a dedicated folder for this exercise, organize its data, scripts, and graphical outputs into separate subfolders, and use relative file paths to import and save files.

You will use a provided iris dataset to practice importing data, creating a filtered dataset, saving the filtered data, and saving a plot. Finally, you will create a ZIP file of your Lab 3D folder. This ZIP file will allow you to submit the organized files you created during the exercise.

```{r}
# Check the working directory
getwd()

# Import the provided iris dataset
iris_data <- read.csv(
  "Lab3D_Organizing/Data/iris_data.csv"
)

# Examine the imported data
head(iris_data)

# Extract observations for the versicolor species
versicolor_data <- subset(
  iris_data,
  Species == "versicolor"
)

print(versicolor_data)

# Save the filtered dataset to the Data folder
write.csv(
  versicolor_data,
  file = "Lab3D_Organizing/Data/versicolor_data.csv",
  row.names = FALSE
)

# Plot sepal length and sepal width for Iris versicolor
plot(
  versicolor_data$Sepal.Length,
  versicolor_data$Sepal.Width,
  xlab = "Sepal Length (cm)",
  ylab = "Sepal Width (cm)",
  pch = 19
)

```