---
title: "Salve_Athena_Lab2-1"
author: "Athena Salve"
date: "2026-09-06"
output: html_document
---

### Student Information

Name: Athena Nicole R. Salve

Section: B2

Date: September 2, 2026

GitHub Repository URL: <https://github.com/arsalve28/BIO118>

# 2A Data Types and Operators

## Introduction

R stores information in different data types, which determine how values are represented and how R can operate on them. Common data types used in biological data analysis include character, numeric, integer, and logical values. Operators in R allow us to perform calculations, compare observations, and construct logical conditions. These operations form the foundation for later programming tasks such as subsetting data, using conditional statements, and automating analyses. In this laboratory exercise, you will work with biological and environmental examples to practice creating and manipulating data types and using operators in R.

### Practice

#### 2.3.1 Monitoring Species Population in Different Habitats

```{r}
habitat_A_population <- 320
habitat_B_population <- 275
total_population <- habitat_A_population + habitat_B_population
is_habitat_A_larger <- habitat_A_population > habitat_B_population
population_difference <- habitat_A_population - habitat_B_population
print(paste("Habitat A Population:", habitat_A_population))
print(paste("Habitat B Population:", habitat_B_population))
print(paste("Total Population:", total_population))
print(paste("Is Habitat A larger than habitat B?", is_habitat_A_larger))
print(paste("Population Difference:", population_difference))
```

#### 2.3.2 Analyzing Environmental Data for Climate Change Impact

```{r}
current_temperature <- 25
predicted_increase <- 3.8
future_temperature <- current_temperature + predicted_increase
current_temp_fahrenheit <- (current_temperature * 9/5) + 32
future_temp_fahrenheit <- (future_temperature * 9/5) + 32
exceeds_threshold <- future_temperature > 5
print(paste("Current Temperature:", current_temperature, "°C"))
print(paste("Predicted Increase:", predicted_increase, "°C"))
print(paste("Future Temperature:", future_temperature, "°C"))
print(paste("Current Temperature in Fahrenheit:", 
            current_temp_fahrenheit, "°F"))
print(paste("Future Temperature in Fahrenheit:", 
            future_temp_fahrenheit, "°F"))
print(paste("Does the future temperature exceeds threshold?", 
            exceeds_threshold))
```

#### 2.3.3 Evaluating Pollutant Levels in a River

```{r}
chemical_X_concentration <- 15.3
chemical_Y_concentration <- 9.8
is_chemical_X_safe <- chemical_X_concentration < 10
is_chemical_Y_safe <- chemical_Y_concentration < 10
are_both_safe <- is_chemical_X_safe & is_chemical_Y_safe
print(paste("Chemical X Concentration:", 
            chemical_X_concentration, "ppm"))
print(paste("Chemical Y Concentration:", 
            chemical_Y_concentration, "ppm"))
print(paste("Is Chemical X Safe?", is_chemical_X_safe))
print(paste("Is Chemical Y Safe?", is_chemical_Y_safe))
print(paste("Are Both Chemicals Safe?", are_both_safe))
```

# 2B Understanding Data Structures

## Introduction

Biological datasets can contain different types of information, including species names, measurements, sampling locations, environmental conditions, and categorical observations. R provides different data structures for organizing these values. The most commonly encountered structures in biological data analysis include:

- vectors- one-dimensional collections of values of the same data type;

- matrices- two-dimensional structures consisting of rows and columns, in which all elements have the same basic data type;

- data frames- two-dimensional tabular structures in which different columns can contain different data types;

- lists- collections that can contain different types and structures of objects;

- factors- structures designed to represent categorical variables.

Choosing an appropriate data structure is important because it determines how data can be stored, accessed, and manipulated.

### Practice

#### 2.2.1 Working with DNA Sequences

```{r}
#install stringr package
library(stringr)

#Character Vector Named dna-sequences
dna_sequences <- c("ATGCGTACGTAGCTAGCGT", 
                   "ATGCGTACGTAGTTAGCGT", 
                   "ATGCGTACGTAGATAGCGT")

#Length of Each DNA Sequence
print("Length of each DNA sequence:")
nchar(dna_sequences)

#Number of Occurrences of the Nucleotide "A" in Each Sequence
print("Number of occurrences of nucleotide A in each sequence:")
str_count(dna_sequences, "A")
```

#### 2.2.2 Monitoring Species Population in Different Habitats

```{r}
#Create Data Frame named biodiversity_data
biodiversity_data <- data.frame(
  Species = c("Oak", "Pine", "Maple", "Birch","Spruce", "Cedar",
              "Ash", "Fir", "Beech"),
  Count = c(120, 85, 100, 60, 150, 75, 55, 95, 110),
  Habitat = c("Forest A", "Forest A", "Forest A", 
              "Forest B", "Forest B", "Forest B", 
              "Forest C", "Forest C", "Forest C")
  )

#Total Number of Individuals Observed in Each Habitat
print("Total number of individuals observed in each habitat:")
tapply(biodiversity_data$Count, biodiversity_data$Habitat, sum)

#Subset Data Frame to Find All Species with a Count Less than 60
low_count_species <- subset(biodiversity_data, Count < 60)
print("Species with count less than 60:")
print(low_count_species)
```

#### Monitoring Temperature Changes

```{r}
#Matrix Named monthly_temps 
monthly_temps <- matrix(
  c(
    2.5, 3.1, 6.2, 10.5, 15.3, 20.1,
    22.4, 21.9, 18.3, 13.8, 7.5, 3.9,
    2.7, 3.5, 6.8, 11.0, 16.0, 21.0,
    23.1, 22.5, 19.0, 14.5, 8.0, 4.1
  ),
  nrow = 12,
  ncol = 2,
)

rownames(monthly_temps) <- c("January", "February", "March", 
                             "April", "May", "June", "July", 
                             "August","September", "October", 
                             "November", "December")
colnames(monthly_temps) <- c("Year_1", "Year_2")

# Calculate the average temperature for each year
avg_temp_per_year <- colMeans(monthly_temps)
print("Average temperature for each year:")
print(avg_temp_per_year)

is_year_2_higher <- avg_temp_per_year["Year_2"] > 
  avg_temp_per_year["Year_1"]
print(paste("Did Year 2 have a higher average temperature?", 
            is_year_2_higher)
      )

# Subset to retrieve temperatures for the summer months
temps_for_summer_months <- monthly_temps[c("June", "July", "August"), ]
print("Temperatures for the summer months:")
print(temps_for_summer_months)
```
