---
title: "Salve_Athena_Lab2-2"
output: html_document
---

### Student Information

Name: Athena Nicole R. Salve

Section: B2

Date: September 2, 2026

GitHub Repository URL: <https://github.com/arsalve28/BIO118>

# 2C Using Control Structure

## Introduction

Control structures in R are fundamental programming tools that allow you to control the flow of your code based on certain conditions and repetitive tasks. These structures include conditional statements like if, else, and loops such as for and while that are crucial for automating repetitive operations.

In biological and environmental science research, control structures help you efficiently analyze large datasets, perform repetitive calculations, and make data-driven decisions based on specific criteria. Mastering control structures is essential for writing robust and efficient R scripts that can handle a variety of data analysis tasks.

The objective of this exercise is to help you understand and apply control structures in R. By the end of this exercise, you will be able to use these control structures to analyze and manipulate datasets relevant to your field.

### Practice

#### 2.4.1 Analyzing Gene Expression Data

```{r}
gene_expression <- data.frame(
  Gene = c("GeneA", "GeneB", "GeneC", "GeneD", "GeneE",
           "GeneA", "GeneB", "GeneC", "GeneD", "GeneE",
           "GeneA", "GeneB", "GeneC", "GeneD", "GeneE"),
  
  Expression_Level = c(15.3, 10.8, 12.5, 18.1, 11.7,
                       20.2, 17.6, 14.3, 19.5, 16.2,
                       13.8, 21.1, 19.7, 22.5, 20.3),
  
  Tissue = c("Tissue1", "Tissue1", "Tissue1", "Tissue1", "Tissue1",
             "Tissue2", "Tissue2", "Tissue2", "Tissue2", "Tissue2",
             "Tissue3", "Tissue3", "Tissue3", "Tissue3", "Tissue3")
)

#View data frame
gene_expression

#For loop and If statement to identify genes with an expression_level above 20
for (i in 1:15) {
  if (gene_expression [i, 2] > 20) {
    print (paste(gene_expression [i, 1],
                 gene_expression [i, 3]))
  }
}

#While loop to find the first gene with expression level below 15

#Select tissue 2
tissue_2 <- gene_expression [6:10,]

#Start with the first gene
i <- 1

#Use while loop until expression level drops below 15
while (tissue_2 [i, 2] >= 15) {
  print(tissue_2 [i, ])
  i <- i + 1
}
```

#### 2.4.2 Monitoring Species Populations

```{r}
#Create a vector named species_population
species_populations <- c(150, 45, 75, 200, 30)

#For loop and if-else statement to classify species
for (i in 1:5) {
  if (species_populations[i] < 50) {
    print(paste("Species", i, "Endangered"))
  } else {
    print(paste("Species", i, "Stable"))
  }
}

#Nested for loops to simulate population growth for each species
species <- c("Species1", "Species2", "Species3", "Species4", "Species5")

for (i in 1:5) {
  
  population <- species_populations[i]
  
  for (year in 1:3) {
    population <- population * 1.05
    print(paste(species[i], "Year", year, "Population:", round(population, 2)))
  }
}
```

#### 2.4.3 Monitoring Water Quality

```{r}
#create a vector named pollution_levels
pollution_levels <- c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100)

#for loop, if statement, and next to skip measurements below 50
for (i in 1:11) {
  if (pollution_levels[i] < 50) {
    next
  }
  
  if (pollution_levels[i] > 50) {
    print(paste("Warning:", pollution_levels[i], "ppm"))
  }
}

#repeat loop to monitor pollution levels until critical level is detected
i <- 1

repeat {
  print(paste("Current pollution level:", pollution_levels[i], "ppm"))
  
  if (pollution_levels[i] >= 90) {
    print(paste("Critical warning:", pollution_levels[i], "ppm"))
    break
  }
  
  i <- i + 1
}
```

# 2D User-Defined Functions

A function is a self-contained block of code designed to perform a specific task. They are used to simplify code, avoid repetition, and enhance readability. Functions take inputs (known as arguments), perform operations on those inputs, and return outputs. In biological and environmental research, user-defined functions can help automate data analysis processes, perform calculations on large datasets, and apply complex algorithms to solve specific scientific problems.

R provides many built-in functions, such as mean(), sum(), and length(). However, biological research often requires calculations or procedures that are specific to a particular study. Functions are powerful tools in R that allow you to encapsulate logic, perform calculations, and manipulate data efficiently.

### Practice

#### 2.2.1 DNA Transcription

```{r}
#Create a function for DNA transcription
transcribe_dna_to_rna <- function(dna_sequence) {
  gsub("T", "U", dna_sequence)
}

#Test the function
dna_sequence <- "ATGCGTAGCTAGCTAGCGT"
transcribe_dna_to_rna(dna_sequence)
```

#### 2.2.2 Predicting Future Population Size

```{r}
#Define the function to predict population size
predict_population <- function(current_population, growth_rate, years) {
  future_population <- current_population * (1 + growth_rate)^years
  return(future_population)
}

#Test the function
population_result <- predict_population(1000, 0.05, 10)
print(population_result)
```

#### 2.2.3 Predicting Future Population Size

```{r}
#Define the function to simulate decay
simulate_decay <- function(initial_concentration, decay_rate) {
  concentration <- initial_concentration
  steps <- 0
  
  while (concentration >= 1) {
    concentration <- concentration * (1 - decay_rate)
    steps <- steps + 1
  }
  
  return(steps)
}

# Test the function
decay_steps <- simulate_decay(50, 0.10)
print(paste("Number of steps:", decay_steps))
```
